Notes on conformal manifolds

### 2025-07-16 Draft_MLFlatDirections.tex
Update of the figures.

\documentclass[11pt]{article}

@@

\tikzstyle{decision} = [
rectangle, 
minimum width=3.2cm, 
minimum height=1.cm, 
text centered, 
draw=coral, 
fill=lightcoral,
line width=1pt,
font=\sffamily\small\bfseries,
aspect=1.3
]

@@

		We then perform gradient descent on the points using TensorFlow’s automatic differentiation \cite{tensorflow2015-whitepaper}. The loss function is defined as:
		\begin{equation}
			\mathcal{L} = \sum_{i = 1}^{n_{points}} ||\left.\nabla V(\vec{X}^{\,i})\,\right|_{\vec{y}=0}||^2
		\end{equation}
		where $\vec{X}^{\,i} = (x^i_1,\dots,x^i_6,\tilde{x}^i_7,\tilde{x}^i_8,\tilde{x}^i_9,x^i_{10},\dots,x^i_{12},\tilde{x}^i_{13})$ and $\vec{y}^{\,i} = (x^i_3,x^i_5,x^i_6,\tilde{x}^i_7,\tilde{x}^i_9,x^i_{11},x^i_{12},\tilde{x}^i_{11})$ denotes the data points. This way we only keep $(x^i_1,x^i_2,x^i_4,\tilde{x}^i_8,x^i_{10})$ alive after we have taken the derivative. As already mentionned above, we use the analytic formula for $\nabla V$, but we could have perform the gradient descent by fully using automatic differentiation.

@@

		\begin{figure}
			\centering
			\includegraphics[scale=0.35]{Figures/log_grad_V_hist.png}
			\quad
			\includegraphics[scale=0.35]{Figures/log_Vp4_hist.png}
			\caption{Left panel: Histogram of the logarithm of the norm of the gradient for each point. Right panel: Histogram of the distance of the value of the potential for each point.}
			\label{fig:Analyze_V_points}
		\end{figure}

@@

		\begin{figure}[h!]
			\centering
			\includegraphics[scale = 0.5]{Figures/Loss_grad_des_1.png}
			\caption{Evolution of the loss function during gradient descent and corresponding learning rate schedule, noted $\alpha$ on the left y-axis, both plotted on a logarithmic scale. Dashed lines indicate epochs at which the optimizer was reinitialized.}
			\label{fig:loss_grad_des}
		\end{figure}

		As a first visualization, we present a triangular plot of the data in Fig. \ref{triangular_plot_1_2_4_8_10}. This figure shows all possible 2D projections of the data, along with the 1D histograms of each coordinate after gradient descent.
		\begin{figure}[h!]
			\centering
			\includegraphics[scale = 0.3]{Figures/triangular_plot_1_2_4_8_10.png} %names need to be changed.
			\caption{Triangular plot showing $2d$ projections and $1d$ histograms of the data after the gradient descent}
			\label{triangular_plot_1_2_4_8_10}
		\end{figure}
		We can make a few comments on those plots. First, we observe non-trivial correlations in the data. Some of these may result from larger basins of attraction, such as in the $x_1/x_2$ plot. However, the structures seen in the $x_4/\tilde{x}_8$ or $x_4/x_{10}$ plots likely reflect genuine features of the manifold, and will show latter that this is indeed the case. We also note that all directions appear to be well populated. 
		%Furthermore, we can notice that a large majority of the points converge to the origin $\vec{0}$. This is not surprising, as this just mean that the round sphere solution has a larger basin of attraction, which sounds reasonable. We do not however attempt to shift the points along flat directions to depopulate the origin or repopulate other regions.
		%\paragraph{Local PCA: graphs of number of patches with given local dimension, and with varying size of local patch, discuss optimal patch size (we want at least few points in each direction)}
		\subsubsection{Determining the dimension}
		Once the gradient descent has been completed and the flat directions sampled, the next step is to identify the structure of the underlying manifold. Our goal is to eventually obtain an analytical expression, not just a numerical description. Before applying symbolic regression to search for such an expression, we can first perform some exploratory analyses to better understand the data. Specifically, we aim to determine the dimensionality of the manifold and wether it consists of a single connected component or multiple disjoint components (e.g., two intersecting hyperplanes).
		To this end, we apply a local Principal Component Analysis (PCA). For each point, we identify its $k$ nearest neighbours and perform a PCA on that local neighbourhood. This procedure allows us to determine how many principal directions are needed to explain a given proportion $\epsilon$ of the data variance. In other words, it provides an estimate of the local dimensionality around each point.
		We perform this analysis for several values of $k$, namely $k \in \{5, 10, 20, 50, 100\}$, and we fix $\epsilon = 0.99$. The results are presented in Fig. \ref{local_pca}.

		\begin{figure}[h!]
			\centering
			\includegraphics[scale=0.4]{Figures/Local_PCA.png}
			\caption{Results of the local PCA analysis. The $x$-axis shows the dimension inferred by the algorithm, and the $y$-axis indicates the proportion of points for which that dimension was found. Each curve corresponds to a different value of $k$ in the $k$-nearest neighbours.}
			\label{local_pca}
		\end{figure}

@@

						\draw [arrow] (end_iter) -- node[right, arrow_label] {No} (exploit);
						\draw [arrow] (exploit) -- (output);
						\draw [arrow] (output) -- (stop);

@@

\ce{Keep fig.~\ref{fig:evolutionoftargetreproduction}(b)?}

\begin{figure}[h!]
	\centering
	\begin{adjustbox}{width=20cm,center}
		\begin{minipage}[t]{0.31\paperwidth}
			\centering
			\includegraphics[width=\linewidth]{Figures/Count_beta_adaptative_temp_indiv}
			\caption*{(a)}
		\end{minipage}%
		\hfill
		\begin{minipage}[t]{0.31\paperwidth}
			\centering
			\includegraphics[width=\linewidth]{Figures/Count_good_pol_all}
			\caption*{(b)}
		\end{minipage}%
		\hfill
		\begin{minipage}[t]{0.31\paperwidth}
			\centering
			\includegraphics[width=\linewidth]{Figures/Loss_all_ASMC.png}
			\caption*{(c)}
		\end{minipage}
	\end{adjustbox}
	\caption{(a) Dynamics of the mean number of particles reproducing each polynomial in eq.~\eqref{eq:pols} (left scale) and evolution of the inverse temperature $\beta$ and its $1\sigma$ deviation (right scale). (b) Evolution of the mean number of particles reproducing a target polynomial, along with its $1\sigma$ deviation, for the parameters described in~\eqref{eq:sample10000adaptativetemp}. (c) Evolution of the losses during the runs (in blue) and their average (in yellow).}
	\label{fig:evolutionoftargetreproduction}
\end{figure}

\ce{Tests sur le gradient analytique pour voir si c'est un polynome annulateur.}

>>>>

\documentclass[11pt,a4paper]{article}

@@

\tikzstyle{decision} = [
rectangle, 
minimum width=3.2cm, 
minimum height=1.cm, 
text centered, 
draw=myorange, 
fill=myorange!20,
line width=1pt,
font=\sffamily\small\bfseries,
aspect=1.3
]

@@

		We then perform gradient descent on the points using TensorFlow’s automatic differentiation \cite{tensorflow2015-whitepaper}. The loss function is defined as:
		\begin{equation} \label{eq:lossgraddes}
			\mathcal{L} = \sum_{i = 1}^{n_{points}} ||\left.\nabla V(\vec{X}^{\,i})\,\right|_{\vec{y}=0}||^2
		\end{equation}
		where $\vec{X}^{\,i} = (x^i_1,\dots,x^i_6,\tilde{x}^i_7,\tilde{x}^i_8,\tilde{x}^i_9,x^i_{10},\dots,x^i_{12},\tilde{x}^i_{13})$ and $\vec{y}^{\,i} = (x^i_3,x^i_5,x^i_6,\tilde{x}^i_7,\tilde{x}^i_9,x^i_{11},x^i_{12},\tilde{x}^i_{11})$ denotes the data points. This way we only keep $(x^i_1,x^i_2,x^i_4,\tilde{x}^i_8,x^i_{10})$ alive after we have taken the derivative. As already mentionned above, we use the analytic formula for $\nabla V$, but we could have perform the gradient descent by fully using automatic differentiation.

@@

		\begin{figure}
			\centering
			\subfigure{ \label{fig:log_grad_V}
				\begin{tikzpicture}
			  		\draw (0,0) node (fig1) {\includegraphics[scale=0.7]{Figures/log_grad_V_hist_21_05_2025.pdf}};
			  		\draw ($(fig1.north west) + (0,0)$) node {\small (a)};
			  		\draw (9,0) node (fig2) {\includegraphics[scale=0.7]{Figures/log_Vp4_hist_21_05_2025.pdf}};
			  		\draw ($(fig2.north west) + (0,0)$) node {\small (b)};
			  	\end{tikzpicture}
			}
			\subfigure{\label{fig:log_Vp4}}
			\vspace{-1cm}
			\caption{(a) Histogram of the the norm of the gradient for each point ($x$ axis is in $\log$ scale). (b) Histogram of the distance of the value of the potential to $V(0)$ for each point ($x$ axis is in $\log$ scale).}
			\label{fig:Analyze_V_points}
		\end{figure}

@@

		\begin{figure}[h!]
			\centering
			\includegraphics[scale = 0.75]{Figures/Loss_grad_des_21_05_2025.pdf}
			\caption{Evolution of the loss function~\eqref{eq:lossgraddes} during gradient descent and corresponding learning rate schedule, noted $\alpha$ on the left y-axis, both plotted on a logarithmic scale. Dashed lines indicate epochs at which the optimizer was reinitialized.}
			\label{fig:loss_grad_des}
		\end{figure}

		As a first visualization, we present a triangular plot of the data in Fig. \ref{triangular_plot_1_2_4_8_10}. This figure shows all possible 2D projections of the data, along with the 1D histograms of each coordinate after gradient descent.
		\begin{figure}[h!]
			\centering
			\makebox[\textwidth][c]{\includegraphics[scale = 0.7]{Figures/Triangular_plot_1_2_4_8_10_21_05_2025.pdf}}
			\caption{Triangular plot showing $2d$ projections and $1d$ histograms of the data after the gradient descent}
			\label{triangular_plot_1_2_4_8_10}
		\end{figure}
		We can make a few comments on those plots. First, we observe non-trivial correlations in the data. Some of these may result from larger basins of attraction, such as in the $x_1/x_2$ plot. However, the structures seen in the $x_4/\tilde{x}_8$ or $x_4/x_{10}$ plots likely reflect genuine features of the manifold, and will show latter that this is indeed the case. We also note that all directions appear to be well populated. 
		%Furthermore, we can notice that a large majority of the points converge to the origin $\vec{0}$. This is not surprising, as this just mean that the round sphere solution has a larger basin of attraction, which sounds reasonable. We do not however attempt to shift the points along flat directions to depopulate the origin or repopulate other regions.
		%\paragraph{Local PCA: graphs of number of patches with given local dimension, and with varying size of local patch, discuss optimal patch size (we want at least few points in each direction)}
		\subsubsection{Determining the dimension}
		Once the gradient descent has been completed and the flat directions sampled, the next step is to identify the structure of the underlying manifold. Our goal is to eventually obtain an analytical expression, not just a numerical description. Before applying symbolic regression to search for such an expression, we can first perform some exploratory analyses to better understand the data. Specifically, we aim to determine the dimensionality of the manifold and wether it consists of a single connected component or multiple disjoint components (e.g., two intersecting hyperplanes).
		To this end, we apply a local Principal Component Analysis (PCA). For each point, we identify its $k$ nearest neighbours and perform a PCA on that local neighbourhood. This procedure allows us to determine how many principal directions are needed to explain a given proportion $\epsilon$ of the data variance. In other words, it provides an estimate of the local dimensionality around each point.
		We perform this analysis for several values of $k$, namely $k \in \{5, 10, 20, 50, 100\}$, and we fix $\epsilon = 0.99$. The results are presented in fig.~\ref{local_pca}.

		\begin{figure}[h!]
			\centering
			\includegraphics[scale=0.75]{Figures/Local_PCA_21_05_2025.pdf}
			\caption{Results of the local PCA analysis. The $x$-axis shows the dimension inferred by the algorithm, and the $y$-axis indicates the proportion of points for which that dimension was found. Each curve corresponds to a different value of $k$ in the $k$-nearest neighbours.}
			\label{local_pca}
		\end{figure}

@@

						\draw [arrow] (end_iter) -- node[right, arrow_label] {Yes} (exploit);
						\draw [arrow] (exploit) -- (output);
						\draw [arrow] (output) -- (stop);

@@

\begin{figure}
	\centering
	\subfigure{ \label{fig:count_pol_all_beta}
		\begin{tikzpicture}
	  		\draw (0,0) node (fig1) {\includegraphics[scale=0.7]{Figures/Count_pol_all_beta_1000runs_13.pdf}};
	  		\draw ($(fig1.north west) + (0,0)$) node {\small (a)};
	  		\draw (8.25,0) node (fig2) {\includegraphics[scale=0.7]{Figures/Count_pol_indiv_1000runs_13.pdf.pdf}};
	  		\draw ($(fig2.north west) + (0,0)$) node {\small (b)};
	  	\end{tikzpicture}
	}
	\subfigure{\label{fig:count_pol_indiv}}
	\vspace{-1cm}
	\caption{(a) Evolution of the mean number of particles reproducing a target polynomial, along with its $1\sigma$ deviation, for the parameters described in~\eqref{eq:sample10000adaptativetemp} (left scale) and evolution of the inverse temperature $\beta$ and its $1\sigma$ deviation (right scale). (b) Dynamics of the mean number of particles reproducing each polynomial in eq.~\eqref{eq:pols} (the scale of the $y$ axis is logarithmic).}
	\label{fig:evolutionoftargetreproduction}
\end{figure}

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.75]{Figures/Loss_all_ASMC_1000runs_13.pdf}
	\caption{Evolution of the losses during the runs (in blue) and their average (in yellow).}
	\label{fig:lossallASMC}
\end{figure}

\ce{Tests sur le gradient analytique pour voir si c'est un polynome annulateur. {\bfseries Say more. Add legend?}}
